Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-05-16 Paper/old/paper.tex

%Can you briefly identify how the fourth item differs from the first in the next para? Maybe put them together as alternative mechanisms?

In some problems of protein science, we want to know the relative stability of a protein's conformation $A$ compared to
another of its conformations $B$. We call this the \emph{difference free energy}. For example, in allosteric mechanisms,
a protein often adopts one conformation when a ligand is bound and another conformation when no ligand is
bound~\cite{Elber2007}. To understand the mechanism requires knowledge of the relative free energies of the different
protein conformations.  Another example is in protein structure prediction, as in the Critical Assessment of
Structure Prediction (CASP) competition~\cite{Moult2011}. If you have used a computational model to predict
two putative native structures, $A$ and $B$, you want to compute which has the lower free energy, in order to know which
is more native-like. Third, you may want to know if mutating a few amino acids in a protein could
cause a protein to switch from one stable conformation to another, because that could have important consequences for
biological mechanisms and disease. Fourth, sometimes binding a ligand induces a protein to switch from one conformation to
another. A quantitative understanding of such \emph{induced-fits} requires knowledge of the different free
energies of the protein in the two states. In all these cases, it would be useful to have a computational method that can
be used efficiently with atomically detailed physical forcefield models to compute the difference free energy between
protein conformations.

@@

%The phrase "protein data bank identifier" is a bit confusing. I can't tell if it means the abbreviation that identifies the structure, or if you mean the structure. If you mean the experimentally determined structure, then I suggest saying "the structure determined by nmr/x-ray crystallography." I suppose that you will want to give the resolution of the experimental structures somewhere...

@@

The GA and GB sequences are nearly identical, so there are some common features observed for all sequences.
The experimental observations classified the protein into two parts: Amino acids 9--51 are fully structured in both
folds, whereas residues 1--8 and 52--56 are unstructured in the $\alpha$ fold, but form $\beta$-strands in the $\beta$
fold. Most of the amino acid residues in the 
%structured?
region 1--9 have negative per-residue free energies, which means that these
residues favor the $\beta$ structure. Roles of some other important residues in stabilizing either the $\alpha$ or
$\beta$ conformation are summarized in Figure S1.

@@

We discussed how the per-residue free energy can reveal mechanistic detail behind the
conformational preference of chameleon sequences. Here, we applied the same method to
see if the per-residue free energy can help us identify problematic residues in structure predictions. We chose
CASP target T0569 and compared the experimental NMR structure with the best predicted model (GDT-TS=78; predicted by the
"Mufold'' group). The confinement method predicts that the experimental structure is more stable by 20 kcal/mol (Figure
S2).
%The annoying rule is that "side chain" is correct for the noun usage, and side-chain for the adjective.
The confinement method clearly identified the two hydrophobic residues V59 and I61as destabilizing the predicted model
relative to the experimental structure (Figure~\ref{fig:T0569_per_residue} and Figure S3). The side chains of these
hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but are oriented towards
the exterior of the protein, exposing them to solvent, in the model. These residues are part of a beta-sheet in the
experimental structure, but because of their side-chain orientation, the corresponding beta-sheet becomes disordered in
the predicted model (Figure~\ref{fig:T0569_per_residue} and Figure S4). There is also a large difference around K76,
which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure. This suggests that
salt-bridge interactions are too favorable for the combination of forcefield and implicit solvent model we use, which
has been a problem noted in the past \cite{Roe2007}.

@@

We calculated the approximate per-residue free energy as follows. The confinement energy, $\Delta G_{A,A*}$ and $\Delta
G_{B,B*}$ of each residue was calculated numerically as described by Tyka et al. \cite{Tyka2006}. The
internal energy of each residue was calculated with Amber's "decomp" module using the final restrained trajectory. We
call this method approximate as we ignore the entropic contribution from the normal mode or quasi-harmonic analysis.
However,
% the entropic
contribution to the total free energy is much smaller, which allow us to study the details of
the conformational preferences of each residue.

\begin{thebibliography}{99}
%You may want to check your styles for creating the references against the style of the Journal to which you are submitting the paper. Also, you may want to check your abbreviations for some of the journal titles against a list such as the one at: http://library.caltech.edu/reference/abbreviations/.  E.g. J. Chem. Phys. or J Chem Phys...Proc. Natl. Acad. Sci. U.S.A. or Proc. Natl. Acad. Sci. 
\bibitem{Elber2007}
Elber, R. A. Milestoning Study of the Kinetics of an Allosteric Transition: Atomically Detailed Simulations of Deoxy Scapharca
Hemoglobin. Biophysical J., 2007, 92, 85-87.

@@

\bibitem{E2007}
%Is E, W.; an error?
E, W.; Ren, W.; Vanden-Eijnden, E. Simplified and improved string method for computing the minimum energy paths in
barrier-crossing events. J. Chem. Phys. 2007, 126, 164103.

@@

\bibitem{Chipot2007}
Chipot, C.; Shell, M.S.; Pohorille, A. Introduction, in Chipot, C., Pohorille, A., editors. Free Energy
Calculations: Theory and Applications in Chemistry and Biology. Springer Series in Chemical
Physics, vol. 86. Berlin and Heidelberg: Springer; 2007, p. 1–32.

@@

\bibitem{Strajbl2000}
Strajbl, M.; Sham, Y.Y.; Villà, J.; Chu, Z.-T.; Warshel, A. Calculations of Activation Entropies of Chemical Reactions
in Solution. (2000) 104, 4578-4584.

@@

\bibitem{Kryshtafovych2011}
Kryshtafovych, A.; Fidelis, K; and Tramontano, A. Evaluation of model quality predictions in CASP9. Proteins, 2011, 79, 91–106

@@

\bibitem{Zemla2003}
Zemla, A. LGA: a method for finding 3D similarities in protein structures. Nucleic Acids Res 2003, 31, 3370–3374.

%I think that Case2012 is the only bibitem where you have the initials before the last name; and the date in parentheses.
\bibitem{Case2012}
D.A. Case, T.A. Darden, T.E. Cheatham, III, C.L. Simmerling, J. Wang, R.E. Duke, R. Luo, R.C. Walker, W. Zhang, K.M. Merz, B. Roberts, S. Hayik, A. Roitberg, G. Seabra, J. Swails, A.W. Goetz, I. Kolossváry, K.F. Wong, F. Paesani, J. Vanicek, R.M. Wolf, J. Liu, X. Wu, S.R. Brozell, T. Steinbrecher, H. Gohlke, Q. Cai, X. Ye, J. Wang, M.-J. Hsieh, G. Cui, D.R. Roe, D.H. Mathews, M.G. Seetin, R. Salomon-Ferrer, C. Sagui, V. Babin, T. Luchko, S. Gusarov, A. Kovalenko, and P.A. Kollman (2012), Amber12, University of California, San Francisco.

>>>>

%Can you briefly identify how the fourth item differs from the first in the next para? Maybe put them together as alternative mechanisms?
%\Arijit{I have removed the line: Fourth, sometimes binding a ligand induces a protein to switch from one conformation to another.}

In some problems of protein science, we want to know the relative stability of a protein's conformation $A$ compared to
another of its conformations $B$. We call this the \emph{difference free energy}. For example, in allosteric mechanisms,
a protein often adopts one conformation when a ligand is bound and another conformation when no ligand is
bound~\cite{Elber2007}. To understand the mechanism requires knowledge of the relative free energies of the different
protein conformations.  Another example is in protein structure prediction, as in the Critical Assessment of
Structure Prediction (CASP) competition~\cite{Moult2011}. If you have used a computational model to predict
two putative native structures, $A$ and $B$, you want to compute which has the lower free energy, in order to know which
is more native-like. Third, you may want to know if mutating a few amino acids in a protein could
cause a protein to switch from one stable conformation to another, because that could have important consequences for
biological mechanisms and disease. A quantitative understanding of such change of conformation requires knowledge of the different free
energies of the protein in the two states. In all these cases, it would be useful to have a computational method that can
be used efficiently with atomically detailed physical forcefield models to compute the difference free energy between
protein conformations.

@@

%The phrase "protein data bank identifier" is a bit confusing. I can't tell if it means the abbreviation that identifies the structure, or if you mean the structure. If you mean the experimentally determined structure, then I suggest saying "the structure determined by nmr/x-ray crystallography." I suppose that you will want to give the resolution of the experimental structures somewhere...
%\Arijit{ I have not changed anything here ... may be we can just write (pdb id)}

@@

The GA and GB sequences are nearly identical, so there are some common features observed for all sequences.
The experimental observations classified the protein into two parts: Amino acids 9--51 are fully structured in both
folds, whereas residues 1--8 and 52--56 are unstructured in the $\alpha$ fold, but form $\beta$-strands in the $\beta$
fold. Most of the amino acid residues in the 
%structured?
%\Arijit {yes}
region 1--8 have negative per-residue free energies, which means that these
residues favor the $\beta$ structure. Roles of some other important residues in stabilizing either the $\alpha$ or
$\beta$ conformation are summarized in Figure S1.

@@

We discussed how the per-residue free energy can reveal mechanistic detail behind the
conformational preference of chameleon sequences. Here, we applied the same method to
see if the per-residue free energy can help us identify problematic residues in structure predictions. We chose
CASP target T0569 and compared the experimental NMR structure with the best predicted model (GDT-TS=78; predicted by the
"Mufold'' group). The confinement method predicts that the experimental structure is more stable by 20 kcal/mol (Figure
S2).
%The annoying rule is that "side chain" is correct for the noun usage, and side-chain for the adjective.
The confinement method clearly identified the two hydrophobic residues V59 and I61 as destabilizing the predicted model
relative to the experimental structure (Figure~\ref{fig:T0569_per_residue} and Figure S3). The side chains of these
hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but are oriented towards
the exterior of the protein, exposing them to solvent, in the model. These residues are part of a beta-sheet in the
experimental structure, but because of their side-chain orientation, the corresponding beta-sheet becomes disordered in
the predicted model (Figure~\ref{fig:T0569_per_residue} and Figure S4). There is also a large difference around K76,
which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure. This suggests that
salt-bridge interactions are too favorable for the combination of forcefield and implicit solvent model we use, which
has been a problem noted in the past \cite{Roe2007}.

@@

We calculated the approximate per-residue free energy as follows. The confinement energy, $\Delta G_{A,A*}$ and $\Delta
G_{B,B*}$ of each residue was calculated numerically as described by Tyka et al. \cite{Tyka2006}. The
internal energy of each residue was calculated with Amber's "decomp" module using the final restrained trajectory. We
call this method approximate as we ignore the entropic contribution from the normal mode or quasi-harmonic analysis.
However, the entropic 
% the entropic
%\Arijit{ I have inserted "the entropic"}
contribution to the total free energy is much smaller, which allow us to study the details of
the conformational preferences of each residue.

\begin{thebibliography}{99}
%You may want to check your styles for creating the references against the style of the Journal to which you are submitting the paper. Also, you may want to check your abbreviations for some of the journal titles against a list such as the one at: http://library.caltech.edu/reference/abbreviations/.  E.g. J. Chem. Phys. or J Chem Phys...Proc. Natl. Acad. Sci. U.S.A. or Proc. Natl. Acad. Sci.

\bibitem{Elber2007}
Elber, R. A. Milestoning Study of the Kinetics of an Allosteric Transition: Atomically Detailed Simulations of Deoxy Scapharca
Hemoglobin. Biophysical J., 2007, 92, 85-87.

@@

\bibitem{E2007}
%Is E, W.; an error?
%\Arijit{E is the last name}
E, W.; Ren, W.; Vanden-Eijnden, E. Simplified and improved string method for computing the minimum energy paths in
barrier-crossing events. J. Chem. Phys. 2007, 126, 164103.

@@

\bibitem{Chipot2007}
Chipot, C.; Shell, M.S.; Pohorille, A. Introduction, in Chipot, C., Pohorille, A., editors. Free Energy
Calculations: Theory and Applications in Chemistry and Biology. Springer Series in Chemical
Physics, vol. 86. Berlin and Heidelberg: Springer; 2007, p. 1-32.

@@

\bibitem{Strajbl2000}
Strajbl, M.; Sham, Y.Y.; Vill√†, J.; Chu, Z.-T.; Warshel, A. Calculations of Activation Entropies of Chemical Reactions
in Solution. (2000) 104, 4578-4584.

@@

\bibitem{Kryshtafovych2011}
Kryshtafovych, A.; Fidelis, K; and Tramontano, A. Evaluation of model quality predictions in CASP9. Proteins, 2011, 79, 91-106

@@

\bibitem{Zemla2003}
Zemla, A. LGA: a method for finding 3D similarities in protein structures. Nucleic Acids Res 2003, 31, 3370-3374.

%I think that Case2012 is the only bibitem where you have the initials before the last name; and the date in parentheses.
%\Arijit{Done it .... but not sure whether it was required to change}
\bibitem{Case2012}
Case, D.A.; Darden, T.A.; Cheatham, III, T.E.; Simmerling, C.L.; Wang, J.; Duke, R.E.; Luo, R.; Walker, R.C.; Zhang, W.; Merz, K.M.; Roberts, B.; Hayik, S.; 
Roitberg, A.; Seabra, J.; Swails, A.W.; Goetz, A.W.; KolossvA°ry, I.; Wong, K.F.; Paesani, F.; Vanicek, J.; Wolf, R.M.; Liu, J.; Wu, X.; Brozell, S.R.; Steinbrecher, T.;
Gohlke, H.; Cai, Q.; Ye, X.; Wang, J.; Hsieh, M.-J.; Cui, G.;  Roe, D.H.; Mathews, D.H.; Seetin, M.G.; Salomon-Ferrer, R.; Sagui, C.; 
Babin, V.; Luchko, T.; Gusarov, S.; Kovalenko, A.; Kollman, P.A. 2012, Amber12, University of California, San Francisco.
